Studying measurement invariance in factor analysis models via stochastic processes

### 2011-02-09 MIpaper/par_instable.tex
Revised simulation/discussion sections
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@37 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

To employ the LR test for assessing measurement invariance, the data
are typically separated into a certain number of given subgroups where
the parameters are estimated separately (for ease of
  exposition, we assume no parameters are restricted to be equal
  across subgroups). Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.
\readme{Achim: You previously modified the above section so that $\theta$
  consists only of parameters that are ``doubled'' under the
  alternative.  This appears to not allow for situations where
  parameters are restricted to be equal across subgroups.  I am ok
  with this, but many applications of LR tests with these models have
  some parameters restricted to be equal across subgroups.  So I've
  added some detail in parentheses at the beginning of the above
  paragraph.}

@@

\subsection{Results}
% MODIFY RESULTS TO DISCUSS ALL FOUR TESTS
% ALSO NEED 0-SE CONDITION
Full simulation results are presented in Figure~\ref{fig:simres}, and
a subset of the results (for finer distinctions) is displayed in
Table~\ref{tab:simres}.  In describing the results, we largely refer
to the figure.

@@

% Need to modify table
\begin{table*}
\caption{Simulated power and Type I error rates for four test statistics
  across three sample sizes and four magnitudes of measurement
  invariance violations. Abbreviations: CvM = Cram\'{e}r-von Mises
  test; MLM = Maximum Lagrange multiplier test; DM = Double-max test; 
  $\chi^2$ = Chi-square likelihood ratio test.}
\label{tab:sim1}
\begin{tabular}{llcccc} \thickline
 & & \multicolumn{4}{c}{Violation Magnitude (SEs)} \\
$n$ & Test Stat & 0 & 1 & 2 & 3 \\\hline
100 & CvM       & 0.05 & 0.10 & 0.33 & 0.63 \\ 
    & MLM       &      & 0.09 & 0.28 & 0.64 \\
    & DM        & 0.03 & 0.06 & 0.19 & 0.41 \\ 
    & $\chi^2$  & 0.07 & 0.16 & 0.49 & 0.83 \\ \hline
200 & CvM       & 0.05 & 0.19 & 0.64 & 0.96 \\
    & MLM       &      & 0.19 & 0.63 & 0.96 \\ 
    & DM        & 0.03 & 0.12 & 0.44 & 0.88 \\ 
    & $\chi^2$  & 0.06 & 0.27 & 0.76 & 0.99 \\ \hline
500 & CvM       & 0.05 & 0.44 & 0.98 & 1.00 \\ 
    & MLM       &      & 0.41 & 0.99 & 1.00 \\
    & DM        & 0.04 & 0.35 & 0.94 & 1.00 \\ 
    & $\chi^2$  & 0.05 & 0.58 & 1.00 & 1.00 \\\thickline
\end{tabular}
\end{table*}

% Table 2 of Nyblom89 contains critical values that may prove useful.
% Hansen 1992 p. 524 has table of critical values that matches 
%  Nyblom89 within reason.
% See Zeileis05 p. 450-451

@@

\subsection{Use in Practice}
% Simultaneous test of Meredith's various types of invariance
The proposed tests give researchers a new set of tools for studying 
measurement invariance.  For example, they give researchers the
opportunity to simultaneously test for many types of measurement
invariance (see, e.g., \citeNP{Mer93}).  The traditional
steps have involved hypothesizing a specific type of invariance and
then testing for it via a Likelihood Ratio Test, but the tests
proposed here do not require one to hypothesize a specific type of
invariance a priori.  Instead, one could define a type of measurement
invariance after seeing the proposed tests' results.

% Researchers interpreting subgroups violating MI
In addition to simultaneously testing different types of measurement
invariance, the proposed tests allow researchers to easily interpret
the nature of the invariance violation.  This is made possible through
the tests' abilities to estimate $\eta$, the threshold dividing
individuals into subgroups that violate
measurement invariance.  While a single $\eta$ was estimate in this
paper, it is also possible to define 
formal rules for estimating multiple $\eta$ parameters.  We plan to
explore these ideas in the future.

@@

One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods (i.e.,
Likelihood Ratio tests) may suffice.
However, we could use the framework developed here to obtain an
alternative $\chi^2$ statistic testing measurement invariance.
Assume the observations are divided into $K$ categories $I_1, I_2,
\ldots, I_K$.
First, we can sum scores within each category of the
demographic variable via:
\begin{equation}
    \label{eq:catsum}
    \Delta W_n(I_k) = \frac{1}{n_k\sqrt{n}} \sum_{i \in I_k}
    {\bm{I}}(\widehat{\bm{\theta}})^{-1/2} {\bm{\psi}}({\bm{y}}_i, \widehat{{\bm
      \theta}}),
\end{equation}
where $n_k$ is the sample size in category $k$ and $n$ is the overall
sample size.  This results in a $K \times p$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $K \times p$ matrix, summing over the squared entries in
the column to obtain a $\chi^2$ statistic with
$(K-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.

\subsection{Extensions}
The proposed family of tests can be extended in various ways.  First,
it is possible to construct an algorithm that recursively defines
groups of individuals violating measurement invariance with respect to
multiple auxiliary variables.  Such an algorithm is related to 
classification and regression trees \cite{BreFri84,MerSha10,StrMal09},
and related algorithms have been developed for generalized linear
models \cite{ZeiHot08} and Rasch models \cite{KopZei10}.

Relatedly,
S\'{a}nchez \citeyear{San09} describes a general
method for partitioning/segmenting structural equation models within a 
partial least squares framework.  This method involves direct maximization of
the likelihood ratio (i.e., fitting the model for various subgroups based on
$V$ and choosing the subgroups with the largest likelihood ratio),
however, instead of the tests described in this paper.  The tests
described here are highly advantageous because they only require the
model to be estimated once.

>>>>

To employ the LR test for assessing measurement invariance, the data
are typically separated into a certain number of given subgroups where
the parameters are estimated separately (for ease of
  exposition, we assume no parameters are restricted to be equal
  across subgroups). Then, the sum of maximized likelihoods 
from the subgroups are compared with the original maximized full-sample likelihood
in a $\chi^2$~test. More specifically, for the special case of two subgroups,
the alternative $H_1^*$ from (\ref{eq:h1*}) with fixed and prespecified $\nu$ is adopted
and the null hypothesis $H_0$ from (\ref{eq:h0}) reduces to ${\bm \theta}^{(A)} = {\bm \theta}^{(B)}$.
The parameter estimated $\hat {\bm \theta}^{(A)}$ can then be obtained from the
observations $i = 1, \dots, m$, say, for which $v_i \le \nu$. Analogously,
$\hat {\bm \theta}^{(B)}$ is obtained by maximizing the likelihood for the
observations $i = m + 1, \dots, n$, for which $v_i > \nu$. The LR test statistic
for the given threshold $\nu$ is then
\begin{equation} \label{eq:lr}
  \mathit{LR}(\nu) ~=~ -2 \left[
         \ell(\hat {\bm \theta}; x_1, \dots, x_n)
   ~-~ \{\ell(\hat {\bm \theta}^{(A)}; x_1, \dots, x_m)
    +    \ell(\hat {\bm \theta}^{(B)}; x_{m+1}, \dots, x_n)\}
    \right],
\end{equation}
which has an asymptotic $\chi^2$ with degrees of freedom equal to the number
of parameters in~${\bm \theta}$.
\readme{Achim: You previously modified the above section so that $\theta$
  consists only of parameters that are ``doubled'' under the
  alternative.  This appears to not allow for situations where some 
  parameters are restricted to be equal across subgroups.  This should
  work for the current presentation, 
  but many applications of LR tests with these models have
  some parameters restricted to be equal across subgroups.  So I've
  added some detail in parentheses at the beginning of the above
  paragraph.}

@@

\subsection{Results}
Full simulation results are presented in Figure~\ref{fig:simres}, and
a subset of the results (for finer distinctions) is displayed in
Table~\ref{tab:simres}.  In describing the results, we largely refer
to the figure.

@@

\begin{table*}
\caption{Simulated power for three test statistics
  across three sample sizes, eight magnitudes of measurement
  invariance violations, and two subsets of tested
  parameters. Abbreviations: CvM = Cram\'{e}r-von Mises 
  test; Max-LM = Maximum Lagrange multiplier test; DM = Double-max test.}
\label{tab:simres}
\begin{center}
\begin{tabular}{lllllllllll}
  \hline
   & & & \multicolumn{8}{l}{Violation Magnitude (SE)} \\
  $n$ &  npars &  Statistic    &  0 & 0.5 &    1 &  1.5
       &    2 &  2.5  &   3 &  3.5 \\ 
  \hline
  100  &  3    &  DM           &  3.1 &    3.6 &    6.6 &   12.4 &   20.6 &   32.1  &  46.1 &   57.8 \\ 
         &         &  CvM            &  5.0 &    5.7 &   10.6 &   19.1 &   32.1 &   49.2  &  67.2 &   78.8 \\ 
         &         &  max-LM          &  6.2 &    6.2 &    9.8 &   15.0 &   24.7 &   39.7  &  57.0 &   69.2 \\ 
         &  19   &  DM           &  2.5 &    2.8 &    3.1 &    4.9 &    7.1 &   12.4  &  19.6 &   25.3 \\ 
         &         &  CvM            &  4.2 &    4.8 &    5.8 &    8.6 &   13.0 &   18.9  &  28.2 &   37.4 \\ 
         &         &  max-LM          &  6.7 &    6.9 &    7.7 &    9.6 &   13.3 &   17.3  &  23.6 &   30.6 \\ 
  200  &  3    &  DM           &  3.7 &    5.2 &   12.5 &   26.3 &   47.0 &   70.6  &  86.8 &   95.0 \\ 
         &         &  CvM            &  5.3 &    7.9 &   18.6 &   39.1 &   64.3 &   85.8  &  96.2 &   99.0 \\ 
         &         &  max-LM          &  6.0 &    8.0 &   14.5 &   30.2 &   54.7 &   79.4  &  93.2 &   98.5 \\ 
         &  19   &  DM           &  3.2 &    3.6 &    5.7 &   10.8 &   21.9 &   40.2  &  61.1 &   76.5 \\ 
         &         &  CvM            &  4.6 &    5.8 &    8.5 &   15.0 &   29.1 &   46.4  &  65.4 &   80.7 \\ 
         &         &  max-LM          &  5.6 &    6.4 &    8.3 &   13.2 &   22.3 &   38.2  &  57.5 &   74.8 \\ 
  500  &  3    &  DM           &  4.1 &    8.3 &   32.2 &   69.0 &   93.4 &   99.5  & 100.0 &  100.0 \\ 
         &         &  CvM            &  4.7 &   11.8 &   42.2 &   81.3 &   98.0 &  100.0  & 100.0 &  100.0 \\ 
         &         &  max-LM          &  4.6 &    9.8 &   33.9 &   75.5 &   97.1 &  100.0  & 100.0 &  100.0 \\ 
         &  19   &  DM           &  3.6 &    4.4 &   14.0 &   40.1 &   75.3 &   96.0  &  99.7 &  100.0 \\ 
         &         &  CvM            &  5.0 &    6.7 &   17.9 &   43.6 &   77.4 &   95.9  &  99.7 &   99.9 \\ 
         &         &  max-LM          &  4.9 &    7.0 &   14.2 &   36.0 &   71.0 &   94.9  &  99.6 &  100.0 \\ 
   \hline
\end{tabular}
\end{center}
\end{table*}

@@

\subsection{Use in Practice}
% Simultaneous test of Meredith's various types of invariance
The proposed tests give researchers a new set of tools for studying 
measurement invariance.  For example, they give researchers the
flexibility to: (1) simultaneously test all model parameters, yielding
results relevant to many types of measurement
invariance (see, e.g., \citeNP{Mer93}), or (2) test a single subset of
model parameters, potentially leading to improved power to detect a
single type of measurement invariance.  The traditional
steps have involved hypothesizing a specific type of invariance and
then testing for it via a Likelihood Ratio Test, but this is
unnecessary under the proposed framework.

% Researchers interpreting subgroups violating MI
In addition to simultaneously testing different types of measurement
invariance, the proposed tests allow researchers to easily interpret
the nature of the invariance violation.  This is made possible through
the tests' abilities to estimate $\nu$, the threshold dividing
individuals into subgroups that violate
measurement invariance.  While a single $\nu$ was assumed in this
paper, it is also possible to define 
formal rules for estimating multiple $\nu$ parameters.  We plan to
explore these ideas in the future.

@@

One issue that was largely unaddressed in this paper involved the use
of categorical $V$ to study measurement invariance.  In this case,
groups are 
already specified in advance, and so traditional methods (i.e.,
Likelihood Ratio tests) may suffice.
However, we could use the framework developed here to obtain an
alternative $\chi^2$ statistic testing measurement invariance.
Assume the observations are divided into $K$ categories $I_1, I_2,
\ldots, I_K$.
First, we can sum scores within each category of the
demographic variable via:
\begin{equation}
    \label{eq:catsum}
    \Delta W_n(I_k) = \frac{1}{n_k\sqrt{n}} \sum_{i \in I_k}
    {\bm{I}}(\widehat{\bm{\theta}})^{-1/2} {\bm{\psi}}({\bm{y}}_i, \widehat{{\bm
      \theta}}),
\end{equation}
where $n_k$ is the sample size in category $k$ and $n$ is the overall
sample size.  This results in a $K \times p$ matrix, with one entry for
each category-by-model parameter combination.  We can test
a specific model parameter for invariance by focusing on the associated
column of the $K \times p$ matrix, summing over the squared entries in
the column to obtain a $\chi^2$-distributed statistic with
$(K-1)$ degrees of freedom \cite{HjoKon02}.
Alternatively, to simultaneously test multiple parameters, we can sum
the $\chi^2$ statistics and degrees of freedom for the individual
parameters.

\subsection{Extensions}
The proposed family of tests can be extended in various ways.  First,
it is possible to construct an algorithm that recursively defines
groups of individuals violating measurement invariance with respect to
multiple auxiliary variables.  Such an algorithm is related to 
classification and regression trees \cite{BreFri84,MerSha10,StrMal09},
with related algorithms being developed for generalized linear
models \cite{ZeiHot08} and Rasch models \cite{KopZei10}.

Relatedly,
S\'{a}nchez \citeyear{San09} described a general
method for partitioning/segmenting structural equation models within a 
partial least squares framework.  This method involves direct maximization of
the likelihood ratio (i.e., fitting the model for various subgroups
defined by 
$V$ and choosing the subgroups with the largest likelihood ratio),
however, instead of the tests described in this paper.  The tests
described here are highly advantageous because they only require the
model to be estimated once.
